Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2014-04-30 paper.tex
Some starting information on the James isolated boundary condition method. More to be added later.

@@

Analytic solutions to the Poisson equation customarily assume that the potential vanishes at large distances from the region of non-zero density. However, on a finite computational domain it is usually not possible to have the edges of the domain be far enough away that the potential can be taken to be zero there. Solving the Poisson equation therefore requires knowledge of the values of the potential on the edges of the computational domain. In principle, they can be computed by doing a direct sum over the mass distribution inside the domain, where the mass in each zone is treated as a point source:
\begin{equation}
  \Phi_{\text{lmn}} = G\, \sum_{\text{i, j, k}} \frac{G \rho_{\text{ijk}}}{|\mathbf{x}_{\text{lmn}} - \mathbf{x}_{\text{ijk}}|}\, \Delta V_{\text{ijk}}.
\end{equation}
Here (i, j, k) are the indices of cells inside the domain, and (l, m, n) are the indices of boundary locations. $\Delta V$ is the volume of the zone. We have implemented this as an option\footnote{It is controlled with the \texttt{gravity.direct\_sum\_bcs} input parameter.} in CASTRO. If there are $N$ zones per spatial dimension, then there are $6 N^2$ boundary zones, and each boundary zone requires a sum over $N^3$ zones, so the direct computation of the boundary conditions scales as $N^5$.  This method is expensive enough that it is not used for hydrodynamics simulations (though it is useful for comparison to the approximate solutions described below).

@@

%Figure~\ref{fig:2levgrav} shows the sequence of updates for $\phi$ on a 2-level grid.

@@

\bibliographystyle{apj}
\bibliography{ws}

>>>>

% allows Greek symbols to be bold
\usepackage{bm}

@@

Analytic solutions to the Poisson equation customarily assume that the potential vanishes at large distances from the region of non-zero density. However, on a finite computational domain it is usually not possible to have the edges of the domain be far enough away that the potential can be taken to be zero there. Solving the Poisson equation therefore requires knowledge of the values of the potential on the edges of the computational domain. In principle, they can be computed by doing a direct sum over the mass distribution inside the domain, where the mass in each zone is treated as a point source:
\begin{equation}
  \Phi_{\text{lmn}} = G\, \sum_{\text{i, j, k}} \frac{G \rho_{\text{ijk}}}{|\mathbf{x}_{\text{lmn}} - \mathbf{x}_{\text{ijk}}|}\, \Delta V_{\text{ijk}}.\label{direct_sum}
\end{equation}
Here (i, j, k) are the indices of cells inside the domain, and (l, m, n) are the indices of boundary locations. $\Delta V$ is the volume of the zone. We have implemented this as an option\footnote{It is controlled with the \texttt{gravity.direct\_sum\_bcs} input parameter.} in CASTRO. If there are $N$ zones per spatial dimension, then there are $6 N^2$ boundary zones, and each boundary zone requires a sum over $N^3$ zones, so the direct computation of the boundary conditions scales as $N^5$.  This method is expensive enough that it is not used for hydrodynamics simulations (though it is useful for comparison to the approximate solutions described below).

@@

%Figure~\ref{fig:2levgrav} shows the sequence of updates for $\phi$ on a 2-level grid.
A clever way to evaluate Equation \ref{direct_sum} for the boundary values of the potential was described by \cite{james77}. In electrostatics a commonly encountered boundary-value problem is that of a positive charge distribution inside of a grounded conducting box. The potential vanishes on the surface of the box. For this to happen, a collection of negative charges must form on the outer surface of the box, to compensate the charge enclosed inside the box. This can be described by a surface charge density $\sigma$. This construction is useful in determining what the potential would be at the location of the surface when the box is removed. We will describe the potential outside of a density distribution that is non-zero only in some finite region as ``isolated'', and denote it by $\Phi$. This is the quantity we want to solve for. The potential in the case where the surface is grounded and conducting is called the ``homogeneous'' potential, and will be denoted by $\phi$. The homogeneous potential is the sum of the isolated potential and also the potential due to the surface charges $\Psi$:
\begin{equation*}
  \phi = \Phi + \Psi.
\end{equation*}
In order for $\phi$ to vanish at the surface, $\Phi$ and $\Psi$ must be of opposite sign, which makes sense since they are generated by opposite-sign charges. Rearranging this equation for the isolated potential yields
\begin{equation}
  \Phi = \phi - \Psi.
\end{equation}
Therefore a strategy for finding $\Phi$ is to solve the Poisson equation in the case where the boundary conditions are homogeneous and obtain $\phi$, and subtract from it the $\Psi$ that would be generated by the necessary compensating surface charges $\sigma$. All of this can be done for the gravitational version of Poisson's equation. Even though we do not commonly speak of negative gravitational charges, they are simply a mathematical tool for obtaining the correct potential. In general,
\begin{equation}
  \sigma = \frac{1}{4\pi G} {\bm{\nabla}} \phi \cdot \hat{\mathbf{n}},
\end{equation}
where $\hat{\mathbf{n}}$ is the unit vector perpendicular to (and pointing away from) the surface. On the computational domain, using an appropriate discretized representation of the ${\bm{\nabla}}$ operator, we calculate $\sigma$ at every boundary location. We can then find the resulting potential by summing over all of the surface charges:
\begin{equation}
  \Psi_{\text{l,m,n}} = \sum_{\text{l$^\prime$,m$^\prime$,n$^\prime$}} \frac{G\sigma_{\text{l$^\prime$,m$^\prime$,n$^\prime$}}}{\left|\mathbf{x}_{\text{l,m,n}} - \mathbf{x}_{\text{l$^\prime$,m$^\prime$,n$^\prime$}}\right|} dA_{\text{l$^\prime$,m$^\prime$,n$^\prime$}}
\end{equation}

%In general,
%\begin{equation}
%  \sigma = 4\pi G

@@

\bibliographystyle{apj}
\bibliography{refs}
